Backward Population Synthesis: Mapping the Evolutionary History of Gravitational-Wave Progenitors

### 2022-10-25 src/tex/ms.tex
Adding Zenodo record citation.
add in inference references

\newcommand{\flatiron}{\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}}
\newcommand{\stonybrook}{\affiliation{Department of Physics and Astronomy, Stony Brook University, Stony Brook NY 11794, USA}}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{showyourwork}
\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\newcommand{\wf}[1]{\textcolor{cyan}{WF: #1}}
\newcommand{\kb}[1]{\textcolor{pink}{#1}}
% \setlength{\floatsep}{0cm}
% \setlength{\textfloatsep}{0cm}
% \setlength{\belowcaptionskip}{-5pt}
% \setlength{\abovecaptionskip}{-5pt}
% Begin!
\begin{document}
\title{Backward Population Synthesis: Mapping the Evolutionary History of Gravitational-Wave Progenitors}

@@

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of stellar
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties.  Here
we propose a pipeline to simultaneously infer zero-age main sequence properties
and population synthesis parameter settings controlling modeled binary evolution
from individual gravitational wave observations of merging compact binaries. Our
pipeline can efficiently explore the high-dimensional space of population
synthesis settings and progenitor system properties for each system in a catalog
of gravitational wave observations.  We apply our pipeline to observations in
the third LIGO--Virgo Gravitational-Wave Transient Catalog. We showcase the
effectiveness of this pipeline with a detailed study of the progenitor
properties and population synthesis settings\textbf{, which represent 10 degrees
of freedom in total,} that produce mergers like the observed GW150914.  Our
pipeline permits a measurement of the variation of population synthesis
parameter settings with binary properties, if any; we \textbf{show that} the
recent GWTC-3 transient catalog \textbf{could be consistent with a model where
the stable mass transfer efficiency parameter varies with the primary black hole
mass.}  
\end{abstract}

@@

\section{Introduction}
As the detection rate of gravitation waves (GWs) from merging
double-compact-object (DCO) binaries increases with the sensitivity of the
ground-based GW detector network
\citep{Aso2013,LIGOScientificCollaboration2015,Acernese2015,Abbott2018,Buikema2020,Tse2019,Acernese2019,Akutsu2021},
we are beginning to constrain the astrophysical processes which shape the
evolution of GW progenitor populations. One of the most common ways to study
progenitor populations of GW mergers is through population synthesis simulations
of stellar populations from a host of formation environments. In the case of isolated binary star evolution, merging
double compact objects can form from massive binary stars through the
standard stable mass transfer or common envelope channels
\citep[e.g.][]{Belczynski2002, Dominik2012, Belczynski2016, Stevenson2017, Zevin2020, Bavera2021, Broekgaarden2021, VanSon2022}
as well as through chemically homogenous evolution \citep{Mandel2016, Marchant2016, deMink2016}
or with population III stars \citep[e.g.][]{Belczynski2004, Kinugawa2014, Inayoshi2016, Inayoshi2017, Tanikawa2021, Tanikawa2022}.
Merging DCO binaries can also originate from a wide variety of dynamically active environments including
triple (or higher multiple) systems \citep[e.g.][]{Antonini2017, Silsbee2017, Fragione2019, VignaGomez2021},
as well as young stellar clusters \citep[e.g.][]{Ziosi2014, Banerjee2017, DiCarlo2020, Chattopadhyay2022},
globular clusters \citep[e.g.][]{PortegiesZwart2000, OLeary2006, Downing2010, Samsing2014, Rodriguez2015, Rodriguez2016, Askar2017, Rodriguez2019},
nuclear star clusters \citep[][]{Miller2009, Antonini2016}, or a mix of all three \citep[e.g.][]{Mapelli2022}.
Finally, more exotic channels like active galactic nuclei which combine both gravitational and gas interactions
\citep[e.g.][]{McKernan2018, McKernan2020, Secunda2020, Ford2021} or primordial black holes \citep[e.g.][]{Bird2016, AliHaimoud2017}
can produce merging DCOs observable with a ground-based detector network. For a discussion of the relative rates of each
formation environment and channels within, see \citet{Mandel2022} and references therein.

Traditionally, population synthesis studies simulate merging DCO populations
with a Monte-Carlo approach.  Using theoretically motivated distributions of
initial parameters like age, metallicity, component mass, orbital separation,
and eccentricity, initial populations are evolved with fixed astrophysical
assumptions, or population synthesis hyperparameter settings like the stability 
of Roche-overflow mass transfer, common envelope ejection efficiency, or the 
strength of compact-object natal kicks, to produce a
synthetic catalog of DCO mergers which can be compared to parameterized models
derived from observations. This approach relies on several assumptions, both in
the simulations and model parameterizations fitted to the GW data, which are
unlikely to be fully correct.  For example, there is no reason to suppose that
the entire population of DCOs should evolve with the same fixed hyperparameters;
there is no a priori reason the fitted parameterized model should capture the
relevant features of the observed DCO distributions to correspond to the binary
physics of interest; etc. A better approach would be to compare progenitor
parameters and hyperparameter settings directly to DCO observations, treating
the population synthesis model as a mapping from progenitor to merger
parameters; merger parameters can then be mapped into observations using a
gravitational waveform family, and parameter inference proceeds in the usual
way, propagating information back up the mapping chain
\citep[e.g.][]{Veitch2015}.  A similar methodology, though with \emph{fixed}
hyperparameter settings, was advanced in \citet{Andrews2018,Andrews2021}, as we
discuss in more detail below.

%One of the most common skepticisms of population synthesis studies is that they employ simulation sets which often
%cover a small subdomain of the entire parameter space of uncertain physics, or only vary one uncertain process per simulation
%(e.g. common envelope efficiency) while holding all others fixed to some ``default" values.
%On top of this, these values are often not calibrated to data.
Attempting to approach such an improved inference procedure for evolutionary
hyperparameters using traditional en mass Monte Carlo simulations of entire
populations of DCO mergers is computationally infeasible. Even with recent
developments in using emulators to speed up the simulation process
\citep[e.g.][]{Wong2021}, training the emulators still requires a significant
amount of computation to cover a wide range of uncertainty in the hyperparemeter
space. As an example, to emulate a model with $10$ parameters, the simplest way
to construct a training set of simulations is to run simulations on a grid which
varies combinations of uncertain physics. For $10$ uncertain physical processes,
if we consider $2$ variations for each physical process this would still require
$1024$ simulations. Because population synthesis simulations require hundreds to
thousands of cpu hours, training an emulator which spans a high dimensional
uncertainty space remains an impossibility at present.  Thus \textbf{it is
beneficial to infer hyperpameter settings \emph{dynamically}}, generating trial
DCO systems via population synthesis recipes to match a particular observation
one- or few-at-a-time while simultaneously adjusting the hyperparameters.
(Eventually it may be even be possible to use more physically-motivated modeling
of binary physics \citep[e.g.][]{Gallego-Garcia2021} in such a procedure.)

@@

Retaining explicit control over random variables allows us to marginalize over
their contribution, and thus allows us to focus on progenitor parameters and
hyperparameters. The strength and direction of natal kicks for compact objects 
are directly specified in the \texttt{COMPAS} binary population synthesis code 
\citep{Riley2022} and can also be specified in \texttt{COSMIC}, however 
this requires the user to specify the strength and direction of natal kicks as 
inputs through a user-specified pre-processing script.  In this study, we use 
\texttt{COSMIC} in it's default state which does not specify the kick strength 
and direction at run time. %However, in practice most population synthesis codes (including
%\texttt{COSMIC} which we use in this study) define their random variables
%implicitly such that the random variables are drawn within the program instead
%of being passed as arguments by default 
This means that the process of evolving the binary is not fully
deterministic\textbf{, i.e. we only have access to $\bm{\theta'}$ and
$\bm{\lambda}$ but not $\bm{X}$}. Thus, even if the root-finding algorithm
performs perfectly, forward modelling a set of roots does not guarantee that the
simulated population reproduces exactly the set of posterior samples due to
randomness in the evolution of each binary. To assure that the recovered
progenitor and hyperparameters robustly correspond to the posterior in the
GW-observable space, we push forward (or ``reproject'') the recovered
evolutionary parameters to the GW-observable space to check whether the
reprojected posterior agrees with the posterior given by the LVK collaboration.
\textbf{In a symbolic form, this means we generate a new $\bm{X}$ for each pair of $\bm{\theta'}$ and
$\bm{\lambda}$, then compute the new posterior in $\bm{\theta}$.}

% Validity paragraph
%\wf{What if we made this a subsection of ``methods'', and then just one sentence here about the values of the KL divergence for GW150914?}
%To check the validity of the recovered posterior over evolutionary parameters,
%it is essential to check whether it is consistent with the original posterior in
%the observable space. We reproject the recovered posterior to the observable
%space, and compare the KL divergence between the original posterior and the
%reprojected posterior.

@@

\noindent A small KL divergence means the reprojected \texttt{COSMIC} posterior is similar to the original posterior in the
observable space, and is thus a viable channel for that specific event. Otherwise, it either means \texttt{COSMIC}
can only explain part of the posterior or cannot explain the event at all. This is completely expected behavior since
\texttt{COSMIC}, and isolated binary evolution more generally, carries its own assumptions and is expected to
fail in reproducing a subset of the event in GWTC-3. One example of this is GW events with at least one component
with mass above the pair instability supernova mass limit \citep{Woosley2017, Farmer2019}.
The reprojection could also be subject to stochasticity in the evolution of each binary.
To account for this, we reproject the posterior multiple times with different random seeds and
check whether the KL divergence varies significantly. A varying KL divergence means a particular GW event is subject
to randomness in the evolution of the progenitor, and extra caution should be used when interpreting the result.

@@

We apply the method described in section \ref{sec:method} to the first GW event GW150914 \citep{GW150914}.
We use the \texttt{Overall\_posterior} posterior samples for GW150914,
publicly available from the Gravitational wave Open Science Center \citep{LIGOScientific:2019lzm}. 
The parameters involved in the analysis and the range we allow for the inference are tabulated in table \ref{tab:parameters}.
The spin of BHs formed in isolated binaries depends strongly on the angular
momentum transport in massive stars which is still highly uncertain (see however \citet{Fuller2019, Bavera2020}),
therefore we only consider the two component masses in the observable space in both the
root-finding stage and the MCMC stage. For progenitor parameters, we characterize each guess with
five parameters: the two component masses $M_{\rm 1,ZAMS}$ and $M_{\rm 2,ZAMS}$, the orbital period $t_{\rm orb}$,
the eccentricity $e$, and the metallicity $Z$ at ZAMS. Note that the progenitor formation redshift is not
obtained through the root-finding process because the formation redshift does not intrinsically affect
the evolution of the binary. Instead, we compute the time it takes for the binary to evolve from
formation to the merger and add this time to the lookback time when a GW event is observed and limit the total
time to merger to be less than $13.7\,\rm{Gyr}$.
Using the total lookback time, we can compute the redshift of the binary at ZAMS formation following the
Planck 2018 cosmological model \citep{Planck2018}.
The main benefit of post-processing the redshift in this way is that we do not rely on a particular assumption
of star formation rate (SFR) distribution when we solve for the posterior distribution in redshift.
This means we can put the prior on formation redshift (or in general, metallicity-redshfit distribution) in
by reweighting the posterior distribution according to a particular SFR model,
and therefore do not have to rerun the inference every time we change the SFR model. This is especially helpful
since SFR models are still highly uncertain but strongly impact the local merger rates of BBHs \citep[e.g. ][]{Broekgaarden2021}.
For hyperparameters, we choose parameters that strongly affect
binary evolution in \texttt{COSMIC} for massive stars, including the common envelope
efficiency $\alpha$, the fraction of mass accreted onto the accretor during stable mass transfer $f_{\rm acc}$,
the critical mass ratio which determines whether Roche-overflow mass transfer
proceeds stably or produces a common envelope when the donor star is on the Hertzsprung gap
$q_{\rm crit,3}$, and the root-mean-square of the Maxwellian distribution
employed for CO natal kicks, $\sigma$. For systems which begin a common envelope 
with a Hertzsprung-gap donor, we always assume the pessimistic outcome that the two 
stars merge since the donor has likely not formed a strong core-envelope boundary 
\citep{Ivanova2004, Belczynski2008}.

@@

% Paragraph discussing the potential application on the population side.
To illustrate the potential benefit of using our method on the population level,
we perform the same analysis for all events in GWTC-3.
For events announced in GWTC-1, we use \texttt{Overall\_posterior}, \texttt{PublicationSamples},\texttt{IMRPhenomXPHM\_comoving} and \texttt{C01:Mixed} posterior samples, respectively.
All the data are publicly available from the Gravitational wave Open Science Center \citep{LIGOScientific:2019lzm}.
Plotted in figure
\ref{fig:GWTC-3_f_acc_mass} is the posterior density in the $m_{1,{\rm
GW}}-f_{\rm acc}$ space for most of the events in GWTC-3.
Each contour represents
the $68\%$ credible interval of the posterior density for that particular event.
There is a suggestive trend showing that $f_{\rm acc}$ could increase as the mass of
the progenitor increases. Such a trend implies the stable mass transfer
phase of a less massive binary would be preferentially non-conservative, with
more conservative mass transfer in more massive binary systems.

@@

For high-mass binaries, \texttt{COSMIC} struggles to produce events above the
pair instability supernova mass cutoff, so the posterior in the
evolutionary parameter space only corresponds to part of the posterior in the
observable space below this cutoff, and therefore events beyond the lower edge of the PISN mass
gap also have higher KL divergence.

Events with more extreme mass ratios are hard to produce with \texttt{COSMIC},
and therefore less likely to be accurately recovered by our method \citep[e.g.][]{Zevin2020}. This could
be due to our method assuming a single value for $f_{\rm acc}$ and $\alpha$ as
discussed in Section~\ref{sec:result}. Here we impose a cut to eliminate events with KL divergence larger
than 0.1.
This is a heuristic choice to exclude events that are obviously not compatible with
\texttt{COSMIC}.
\textbf{A careful treatment of all events in the catalog and
discussion related to the detailed physical implications of figure
\ref{fig:GWTC-3_f_acc_mass} is beyond the scope of this paper.  We defer a
detailed study of the physics related to the population of GW events to future
work.

Figure \ref{fig:GWTC-3_f_acc_mass} shows that our method could in principle
reveal the correlation between progenitor parameters and hyperparameters on a
population level.
Merger observations for which \texttt{COSMIC} struggles are likely to be very
informative about formation channels and binary evolution physics, and are
therefore likely worthy of close study.  We anticipate that future work will
follow up these events in detail.}

@@

% Rate analysis is now just counting, like what's done in phenomenological model, instead of forward modelling.
Working event-by-event as we do permits post-processing application of
arbitrarily complicated models of star formation and metalliticy evolution by
re-weighting the samples we obtain \citep[e.g.][]{VanSon2022}. Once we have
pulled the GW event posterior back from the observable space to the evolutionary
parameter space, we can apply the same hierarchical inference methods currently
used to infer the compact binary population to our progenitor population.  In
particular, we can use our progenitor population to infer the formation rate of
compact binary population progenitor systems over cosmic time without needing to
make artifical and simplistic assumptions about the delay time distribution or
the metallicity-specific star formation rate
\citep{Vitale2019,Ng2021,VanSon2022}.  Comparing the formation rate of
progenitors to the star formation rate provides an avenue to check our
understanding of binary evolution and the relative rate of other formation
channels.

% Single formation channel and fundamental limitation on that.
While our method allows data-driven exploration of the hyperparameters space for the first time, there are a number of improvements that can be implemented in future studies.
In this study we use only \texttt{COSMIC} as our evolutionary function, which by design cannot explain all the events in GWTC-3.
For example, events with either of the component mass larger than the lower edge of PISN gap such as GW190521 cannot be explained by \texttt{COSMIC}.
Alternative channels such as dynamical formation might be needed to explain some subset of the events in GWTC-3 \citep{Zevin2021}.
As the sensitivity of GW detectors increases, we expect to see more and more events that are unusual in some way.
Therefore, having a self-consistence population synthesis code that contains multiple formation channels is essential to accommodate the growing catalog of GW events.
In this paper, our main focus is to illustrate the concept of "back-propagating" GW-event posterior samples,
highlighting the capability of our method and motivating the benefits of building population synthesis codes that can work seamlessly with our method.
To avoid cluttering of focus, we discuss the physical implications of the results presented in this work under our specific assumptions (i.e. using \texttt{COSMIC} as our evolutionary function) in a future companion paper.

% Randomness paragraph
Due to the implicit definition of random variables in \texttt{COSMIC}, our evolutionary function is stochastic.
This introduces significant inefficiency in our root-finding and sampling algorithm.
The main stochasticity in \texttt{COSMIC} comes from natal kick, which significantly affect the evolutioary pathway of low mass events such as BNS mergers.
The effect of the natal kick is suppressed for heavier mass events due to fallback.
This means events with lighter masses are subject to stochasticity of the function,
where the sampling process for heavier events behaves as if the evolutionary function we use is deterministic.
% Indeed, we see this behavior in a couple of case studies.
% For heavy event such as GW150914, the reprojected posterior agrees with the data posterior.
% For light event such as GW170817, although during the root-finding process the convergence criteria is met,
% the reprojected posterior does not agree with the data posterior.
Due to computational limitation, we only try 1000 different initial guesses per posterior sample in the root-finding process.
This means any posterior sample that has a probability of merging rarer than 1 in 1000 could be missed.
Obviously the problems which comes with the randomness can be alleviated by performing more tries per posterior sample,
but this is not scalable in practice.
On top of limitation in efficiency, some formation channels require explicit control of random variables by construction.
For example, in a dynamical formation scenario such as binaries that form in a globular cluster,
each binary has some probability of undergoing a multi-body encounter with another member in the cluster.
These encounter probability distributions are either studied with direct N-body simulations or semi-analytical methods.
In both cases, each member of the cluster is no longer completely independent of the other members, but coupled through the encounter probability distribution.
By studying the encounter probability distribution, we can infer the properties of the environment which the binary lives in.
This can only be done if we have explicit control over the random variables that characterize the encounter probability distribution.

@@

% Gradient decending in the evolutionary parameter space is much more efficent than rejection sampling.
% Potential pitfall of interpreting this result.
% Need for full autodiff.

>>>>

\newcommand{\flatiron}{\affiliation{Center for Computational Astrophysics,
Flatiron Institute, New York, NY 10010, USA}}
\newcommand{\stonybrook}{\affiliation{Department of Physics and Astronomy, Stony
Brook University, Stony Brook NY 11794, USA}}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{showyourwork}
\definecolor{rb4}{HTML}{27408B}
\newcommand{\kw}[1]{{\color{rb4}[KW: #1 ]}}
\newcommand{\wf}[1]{\textcolor{cyan}{WF: #1}}
\newcommand{\kb}[1]{\textcolor{pink}{#1}}
% \setlength{\floatsep}{0cm} \setlength{\textfloatsep}{0cm}
% \setlength{\belowcaptionskip}{-5pt} \setlength{\abovecaptionskip}{-5pt} Begin!
\begin{document}
\title{Backward Population Synthesis: Mapping the Evolutionary History of Gravitational-Wave Progenitors}

@@

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of stellar
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties.  Here
we propose a pipeline to simultaneously infer zero-age main sequence properties
and population synthesis parameter settings controlling modeled binary evolution
from individual gravitational wave observations of merging compact binaries. Our
pipeline can efficiently explore the high-dimensional space of population
synthesis settings and progenitor system properties for each system in a catalog
of gravitational wave observations.  We apply our pipeline to observations in
the third LIGO Virgo Kagra Gravitational Wave Transient Catalog. We showcase the
effectiveness of this pipeline with a detailed study of the progenitor
properties and population synthesis settings that produce mergers like the
observed GW150914.  Our pipeline permits a measurement of the variation of
population synthesis parameter settings with binary properties, if any; we
present inferences for the recent GWTC-3 transient catalog that suggest that the
stable mass transfer efficiency parameter may vary with primary black hole mass.  
\end{abstract}

@@

\section{Introduction}
As the detection rate of gravitation waves (GWs) from merging
double-compact-object (DCO) binaries increases with the sensitivity of the
ground-based GW detector network
\citep{Aso2013,LIGOScientificCollaboration2015,Acernese2015,Abbott2018,Buikema2020,Tse2019,Acernese2019,Akutsu2021},
we are beginning to constrain the astrophysical processes which shape the
evolution of GW progenitor populations. One of the most common ways to study
progenitor populations of GW mergers is through population synthesis simulations
of stellar populations from a host of formation environments. In the case of
isolated binary star evolution, merging double compact objects can form from
massive binary stars through the standard stable mass transfer or common
envelope channels
\citep[e.g.][]{Belczynski2002, Dominik2012, Belczynski2016, Zevin2020, Bavera2021, Broekgaarden2021, VanSon2022}
as well as through chemically homogenous evolution \citep{Mandel2016,
Marchant2016, deMink2016} or with population III stars
\citep[e.g.][]{Belczynski2004, Kinugawa2014, Inayoshi2016, Inayoshi2017,
Tanikawa2021, Tanikawa2022}. Merging DCO binaries can also originate from a wide
variety of dynamically active environments including triple (or higher multiple)
systems \citep[e.g.][]{Antonini2017, Silsbee2017, Fragione2019, VignaGomez2021},
as well as young stellar clusters \citep[e.g.][]{Ziosi2014, Banerjee2017,
DiCarlo2020, Chattopadhyay2022}, globular clusters
\citep[e.g.][]{PortegiesZwart2000, OLeary2006, Downing2010, Samsing2014,
Rodriguez2015, Rodriguez2016, Askar2017, Rodriguez2019}, nuclear star clusters
\citep[][]{Miller2009, Antonini2016}, or a mix of all three
\citep[e.g.][]{Mapelli2022}. Finally, more exotic channels like active galactic
nuclei which combine both gravitational and gas interactions
\citep[e.g.][]{McKernan2018, McKernan2020, Secunda2020, Ford2021} or primordial
black holes \citep[e.g.][]{Bird2016, AliHaimoud2017} can produce merging DCOs
observable with a ground-based detector network. For a discussion of the
relative rates of each formation environment and channels within, see
\citet{Mandel2022} and references therein.

Traditionally, population synthesis studies simulate DCO populations with a
Monte-Carlo approach. Using theoretically motivated distributions of initial
parameters like age, metallicity, component mass, orbital separation, and
eccentricity, initial populations are evolved with fixed astrophysical
assumptions, or population synthesis hyperparameter settings, to produce a
synthetic catalog of DCO mergers which can be compared to parameterized models
derived from observations. \textbf{This method has been applied extensively to
inference on GW data to inform phenomenological models
\citep[e.g.][]{Fishbach2017, Wysocki2019, Callister2021, Delfavero2021,
Farah2022, Delfavero2022}, population synthesis for a single formation
environment \citep[e.g.][]{Belczynski2016, Stevenson2017b, Taylor2018,
Barrett2018, Mastrogiovanni2022} or contributions from multiple formation
environments \citep[e.g][]{Zevin2017, Stevenson2017, Bouffanais2019, Zevin2021,
Wong2021, Bouffanais2021, Mapelli2022}}. These approaches rely on several
assumptions, both in the simulations and model parameterizations fitted to the
GW data, which are unlikely to be fully correct. For example, there is no reason
to suppose that the entire population of DCOs should evolve with the same fixed
hyperparameters; there is no a priori reason the fitted parameterized model
should capture the relevant features of the observed DCO distributions to
correspond to the binary physics of interest; etc. A better approach would be to
compare progenitor parameters and hyperparameter settings directly to DCO
observations, treating the population synthesis model as a mapping from
progenitor to merger parameters; merger parameters can then be mapped into
observations using a gravitational waveform family, and parameter inference
proceeds in the usual way, propagating information back up the mapping chain
\citep[e.g.][]{Veitch2015}.  A similar methodology, though with \emph{fixed}
hyperparameter settings, was advanced in \citet{Andrews2018,Andrews2021}, as we
discuss in more detail below.

%One of the most common skepticisms of population synthesis studies is that they
%employ simulation sets which often cover a small subdomain of the entire
%parameter space of uncertain physics, or only vary one uncertain process per
%simulation (e.g. common envelope efficiency) while holding all others fixed to
%some ``default" values. On top of this, these values are often not calibrated
%to data.
Attempting to approach such an improved inference procedure for evolutionary
hyperparameters using traditional en mass Monte Carlo simulations of entire
populations of DCO mergers is computationally infeasible. Even with recent
developments in using emulators to speed up the simulation process
\citep[e.g.][]{Wong2021}, training the emulators still requires a significant
amount of computation to cover a wide range of uncertainty in the hyperparemeter
space. As an example, to emulate a model with $10$ parameters, the simplest way
to construct a training set of simulations is to run simulations on a grid which
varies combinations of uncertain physics. For $10$ uncertain physical processes,
if we consider $2$ variations for each physical process this would still require
$1024$ simulations. Because population synthesis simulations require hundreds to
thousands of cpu hours, training an emulator which spans a high dimensional
uncertainty space remains an impossibility at present.  Thus inference of
hyperpameter settings must proceed \emph{dynamically}, generating trial DCO
systems via population synthesis recipes to match a particular observation one-
or few-at-a-time while simultaneously adjusting the hyperparameters. (Eventually
it may be even be possible to use more physically-motivated modeling of binary
physics \citep[e.g.][]{Gallego-Garcia2021} in such a procedure.)

@@

Retaining explicit control over random variables allows us to marginalize over
their contribution, and thus allows us to focus on progenitor parameters and
hyperparameters. However, in practice most population synthesis codes (including
\texttt{COSMIC} which we use in this study) define their random variables
implicitly such that the random variables are drawn within the program instead
of being passed as arguments. This means that the process of evolving the binary
is not fully deterministic. Thus, even if the root-finding algorithm performs
perfectly, forward modelling a set of roots does not guarantee that the
simulated population reproduces exactly the set of posterior samples due to
randomness in the evolution of each binary. To assure that the recovered
progenitor and hyperparameters robustly correspond to the posterior in the
GW-observable space, we push forward (or ``reproject'') the recovered
evolutionary parameters to the GW-observable space to check whether the
reprojected posterior agrees with the posterior given by the LVK collaboration.

% Validity paragraph \wf{What if we made this a subsection of ``methods'', and
%then just one sentence here about the values of the KL divergence for
%GW150914?} To check the validity of the recovered posterior over evolutionary
%parameters, it is essential to check whether it is consistent with the original
%posterior in the observable space. We reproject the recovered posterior to the
%observable space, and compare the KL divergence between the original posterior
%and the reprojected posterior.

@@

\noindent A small KL divergence means the reprojected \texttt{COSMIC} posterior
is similar to the original posterior in the observable space, and is thus a
viable channel for that specific event. Otherwise, it either means
\texttt{COSMIC} can only explain part of the posterior or cannot explain the
event at all. This is completely expected behavior since \texttt{COSMIC}, and
isolated binary evolution more generally, carries its own assumptions and is
expected to fail in reproducing a subset of the event in GWTC-3. One example of
this is GW events with at least one component with mass above the pair
instability supernova mass limit \citep{Woosley2017, Farmer2020}. The
reprojection could also be subject to stochasticity in the evolution of each
binary. To account for this, we reproject the posterior multiple times with
different random seeds and check whether the KL divergence varies significantly.
A varying KL divergence means a particular GW event is subject to randomness in
the evolution of the progenitor, and extra caution should be used when
interpreting the result.

@@

We apply the method described in section \ref{sec:method} to the first GW event
GW150914. The parameters involved in the analysis and the range we allow for the
inference are tabulated in table \ref{tab:parameters}. The spin of BHs formed in
isolated binaries depends strongly on the angular momentum transport in massive
stars which is still highly uncertain (see however \citet{Fuller2019,
Bavera2020}), therefore we only consider the two component masses in the
observable space in both the root-finding stage and the MCMC stage. For
progenitor parameters, we characterize each guess with five parameters: the two
component masses $M_{\rm 1,ZAMS}$ and $M_{\rm 2,ZAMS}$, the orbital period
$t_{\rm orb}$, the eccentricity $e$, and the metallicity $Z$ at ZAMS. Note that
the progenitor formation redshift is not obtained through the root-finding
process because the formation redshift does not intrinsically affect the
evolution of the binary. Instead, we compute the time it takes for the binary to
evolve from formation to the merger and add this time to the lookback time when
a GW event is observed and limit the total time to merger to be less than
$13.7\,\rm{Gyr}$. Using the total lookback time, we can compute the redshift of
the binary at ZAMS formation following the Planck 2018 cosmological model
\citep{Planck2018}. The main benefit of post-processing the redshift in this way
is that we do not rely on a particular assumption of star formation rate (SFR)
distribution when we solve for the posterior distribution in redshift. This
means we can put the prior on formation redshift (or in general,
metallicity-redshfit distribution) in by reweighting the posterior distribution
according to a particular SFR model, and therefore do not have to rerun the
inference every time we change the SFR model. This is especially helpful since
SFR models are still highly uncertain but strongly impact the local merger rates
of BBHs \citep{Broekgaarden2021}. For hyperparameters, we choose parameters that
affect binary evolution in \texttt{COSMIC} most significantly for massive stars,
including the common envelope efficiency $\alpha$, the fraction of mass accreted
onto the accretor during stable mass transfer $f_{\rm acc}$, the critical mass
ratio which determines whether Roche-overflow mass transfer proceeds stably or
produces a common envelope when the donor star is on the Hertzsprung gap $q_{\rm
crit,3}$, and the root-mean-square of the Maxwellian distribution employed for
CO natal kicks, $\sigma$.

@@

% Paragraph discussing the potential application on the population side.
To illustrate the potential benefit of using our method on the population level,
we perform the same analysis for all events in GWTC-3. Plotted in figure
\ref{fig:GWTC-3_f_acc_mass} is the posterior density in the $m_{1,{\rm
GW}}-f_{\rm acc}$ space for most of the events in GWTC-3. Each contour
represents the $68\%$ credible interval of the posterior density for that
particular event. There is a suggestive trend showing that $f_{\rm acc}$ could
increase as the mass of the progenitor increases. Such a trend implies the
stable mass transfer phase of a less massive binary would be preferentially
non-conservative, with more conservative mass transfer in more massive binary
systems.

@@

For high-mass binaries, \texttt{COSMIC} struggles to produce events above the
pair instability supernova mass cutoff $m_{1,{\rm GW}}$, so the posterior in the
evolutionary parameter space only corresponds to part of the posterior in the
observable space below this cutoff, and therefore events beyond the PISN mass
gap also have higher KL divergence.

Events with more extreme mass ratios are hard to produce with \texttt{COSMIC},
and therefore less likely to be accurately recovered by our method
\citep[e.g.][]{Zevin2020}. This could be due to our method assuming a single
value for $f_{\rm acc}$ and $\alpha$ as discussed in Section~\ref{sec:result}.

Merger observations for which \texttt{COSMIC} struggles are likely to be very
informative about formation channels and binary evolution physics, and are
therefore likely worthy of close study.  We anticipate that future work will
follow up these events in detail.

Figure \ref{fig:GWTC-3_f_acc_mass} shows that our method could in principle
reveal the correlation between progenitor parameters and hyperparameters on a
population level. Here we impose a cut to eliminate events with KL divergence
larger than 0.1. This is a heuristic choice to exclude events that are obviously
not compatible with \texttt{COSMIC}. A careful treatment of all events in the
catalog and discussion related to the detailed physical implications of figure
\ref{fig:GWTC-3_f_acc_mass} is beyond the scope of this paper.  We defer a
detailed study of the physics related to the population of GW events to future
work.

@@

% Rate analysis is now just counting, like what's done in phenomenological
% model, instead of forward modelling.
Working event-by-event as we do permits post-processing application of
arbitrarily complicated models of star formation and metalliticy evolution by
re-weighting the samples we obtain \citep[e.g.][]{VanSon2022}. Once we have
pulled the GW event posterior back from the observable space to the evolutionary
parameter space, we can apply the same hierarchical inference methods currently
used to infer the compact binary population to our progenitor population.  In
particular, we can use our progenitor population to infer the formation rate of
compact binary population progenitor systems over cosmic time without needing to
make artifical and simplistic assumptions about the delay time distribution or
the metallicity-specific star formation rate
\citep{Vitale2019,Ng2021,VanSon2022}.  Comparing the formation rate of
progenitors to the star formation rate provides an avenue to check our
understanding of binary evolution and the relative rate of other formation
channels.

% Single formation channel and fundamental limitation on that.
While our method allows data-driven exploration of the hyperparameters space for
the first time, there are a number of improvements that can be implemented in
future studies. In this study we use only \texttt{COSMIC} as our evolutionary
function, which by design cannot explain all the events in GWTC-3. For example,
event with either of the component mass larger than the lower edge of PISN gap
such as GW190521 cannot be explained by \texttt{COSMIC}. Alternative channels
such as dynamical formation might be needed to explain some subset of the events
in GWTC-3 \citep{Zevin2021}. As the sensitivity of GW detectors increases, we
expect to see more and more events that are unusual in some way. Therefore,
having a self-consistence population synthesis code that contains multiple
formation channels is essential to accommodate the growing catalog of GW events.
In this paper, our main focus is to illustrate the concept of "back-propagating"
GW events posterior samples, highlighting the capability of our method and
motivating the community to build the next generation population synthesis code
that can work with our method. To avoid cluttering of focus, we discuss the
physical implication of the result presented in this work under our specific
assumption (i.e. using \texttt{COSMIC} as our evolutionary function) in a
companion paper.

% Randomness paragraph
Due to the implicit definition of random variables in \texttt{COSMIC}, our
evolutionary function is stochastic. This introduces significant inefficiency in
our root-finding and sampling algorithm. The main stochasticity in
\texttt{COSMIC} comes from natal kick, which significantly affect the
evolutioary pathway of low mass events such as BNS mergers. The effect of the
natal kick is suppressed for heavier mass events due to fallback. This means
events with lighter masses are subject to stochasticity of the function, where
the sampling process for heavier events behaves as if the evolutionary function
we use is deterministic.
% Indeed, we see this behavior in a couple of case studies. For heavy event such
% as GW150914, the reprojected posterior agrees with the data posterior. For
% light event such as GW170817, although during the root-finding process the
% convergence criteria is met, the reprojected posterior does not agree with the
% data posterior.
Due to computational limitation, we only try 1000 different initial guesses per
posterior sample in the root-finding process. This means any posterior sample
that has a probability of merging rarer than 1 in 1000 could be missed.
Obviously the problems which comes with the randomness can be alleviated by
performing more tries per posterior sample, but this is not scalable in
practice. On top of limitation in efficiency, some formation channels require
explicit control of random variables by construction. For example, in a
dynamical formation scenario such as binaries that form in a globular cluster,
each binary has some probability of undergoing a multi-body encounter with
another member in the cluster. These encounter probability distributions are
either studied with direct N-body simulations or semi-analytical methods. In
both cases, each member of the cluster is no longer completely independent of
the other members, but coupled through the encounter probability distribution.
By studying the encounter probability distribution, we can infer the properties
of the environment which the binary lives in. This can only be done if we have
explicit control over the random variables that characterize the encounter
probability distribution.

@@

% Gradient decending in the evolutionary parameter space is much more efficent
% than rejection sampling. Potential pitfall of interpreting this result. Need
% for full autodiff.
